Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2022-01-19 analysis/paper/paper.Rmd
Minor adjustment following peer review

The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` variables representing different debitage and tool types. The artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, for example by impacting accessibility by means of watercraft, the relatively constrained geographical distribution of the sites hopefully counteracts some environmentally given sources of variation.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @eigeland2015; @fuglestvedt2007; @mansrud2016]. These studies are, however, based on non-quantitative and less formal methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to less formal and subjectively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)). This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.

@@

```{r var-contrib, fig.cap = "Contribution of variables to the first two components of the PCA. The dotted red line indicates the expected contribution from each variable given a uniform distribution of impact.", fig.height = 3}
# The plots below investigates the contribution of the variables to the
# components. The code has in large part been taken from 
# Bicho & Cascalheira (2020).
var <- get_pca_var(pcalithics)
dim1 <- fviz_contrib(pcalithics, choice = "var", axes = 1,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 1")
dim2 <- fviz_contrib(pcalithics, choice = "var", axes = 2, 
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 2")
dim3 <- fviz_contrib(pcalithics, choice = "var", axes = 3,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 3")
dim4 <- fviz_contrib(pcalithics, choice = "var", axes = 4,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 4")

# Patchwork for arranging plots
(dim1 | dim2) 
#(dim3 | dim4)
```

>>>>

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))`debitage and tool types. These represent artefact categories that have been used consistently across the reports. Consequently, sub-categories that have only been used in the classification of some inventories have been omitted. This for example pertains to what blanks have been used for the production of tools, and thus distinctions such as scrapers produced from flakes or blades, which has only been noted in some of the reports. Furthermore, the artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, for example by impacting accessibility by means of watercraft, the relatively constrained geographical distribution of the sites hopefully counteracts some environmentally given sources of variation.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @eigeland2015; @fuglestvedt2007; @mansrud2016]. These studies are, however, based on non-quantitative or uni- and bivariate methods, leaving the weighting of the many variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)). This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch which have been combined into the category 'small flint tools'. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.

@@

```{r var-contrib, fig.cap = "Contribution of variables to the first two components of the PCA. The dotted red line indicates the expected contribution from each variable given a uniform distribution of impact.", fig.height = 3}
# The plots below investigates the contribution of the variables to the
# components. The code has in large part been taken from 
# Bicho & Cascalheira (2020).
var <- get_pca_var(pcalithics)
dim1 <- fviz_contrib(pcalithics, choice = "var", axes = 1,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 1")
dim2 <- fviz_contrib(pcalithics, choice = "var", axes = 2, 
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 2")

# Patchwork for arranging plots
(dim1 | dim2) 
```
